Deciphering enhancer rules underlying early human neuronal and neural crest development.

### 2025-04-01 manuscript.tex
Add supplement.

states were obtained from all three cell lines (\textbf{FIGURE SUPPLEMENT}) demonstrating the robustness of the

@@

(\textbf{TABLE SUPPLEMENT}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,

@@

the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.892
across identical states (\textbf{FIGURE SUPPLEMENT}).\par

% !!! calculate this corr coefficient for topic based states !!!

@@

area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{FIGURE SUPPLEMENT})), facial mesenchyme

@@

(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{FIGURE SUPPLEMENT})). Of note, a fraction of neural tube

@@

(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{SUPPLEMENTAL TABLE}) and cluster
26 which is embryo specific (CG-rich motif; \textbf{SUPPLEMENTAL TABLE}). Most clusters (\textbf{Fig. 2f-g} and
\textbf{SUPPLEMENTAL FIGURE}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.

@@

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{SUPPLEMENTAL
FIGURE}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both

@@

migratory neural crest cell type of the organoid (\textbf{SUPPLEMENTAL FIGURE}). Both models predict this region
to be specific to the migratory neural crest class (\textbf{SUPPLEMENTAL FIGURE}) and nucleotide importance scores

@@

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities (SUPPLEMENTAL FIG). Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and SUPPLEMENTAL FIG). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and SUPPLEMENTAL FIG). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and SUPPLEMENTAL FIG). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i} and SUPPLEMENTAL FIG). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOX instance in a general FOXA2 target region the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOX instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}. Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}. Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and SUPPLEMENTAL FIG). In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs. The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig. 4d-e} and SUPPLEMENTAL FIG).

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a} and SUPPLEMENTAL FIG). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and SUPPLEMENTAL FIG). The latter is consistent with a recent study where it has been
shown that TWIST1 and homeodomain TFs cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

@@

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that we deem explainable by either of the two models. Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (SUPPLEMENTAL FIG). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and SUPPLEMENTAL FIG). Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and SUPPLEMENTAL FIG). Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig. 6d} and SUPPLEMENTAL FIG) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}. Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and SUPPLEMENTAL FIG). Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig. 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and SUPPLEMENTAL FIG).\par

@@

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (SUPPLEMENT FIG). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par

@@

>>>>

states were obtained from all three cell lines (\textbf{Supplementary Fig. S1a-b}) demonstrating the robustness of the

@@

(\textbf{Supplementary Table S1}). In both systems we could identify neuronal progenitor cells, early differentiating neurons,

@@

the topic) across the organoid and embryo (\textbf{Fig. 1i}) with an average Pearson correlation coefficient of 0.83
across identical states (\textbf{Supplementary Fig. S1c-d}).\par

@@

area under the precision-recall curve (auPR) of 0.21 resp. 0.29 (\textbf{Supplementary Fig. S2a-b})), facial mesenchyme

@@

(\textbf{Fig. 2c}; and auPR of 0.17 for both models (\textbf{Supplementary Fig. S2a-b})). Of note, a fraction of neural tube

@@

(\textbf{Fig. 2e}) regions with the exception of cluster 13 which is organoid specific (AP1 motif; \textbf{Supplementary Table S2}) and cluster
26 which is embryo specific (CG-rich motif; \textbf{Supplementary Table S2}). Most clusters (\textbf{Fig. 2f-g} and
\textbf{Supplementary Table S2}) contain at least one motif learned \textit{de novo} by the organoid and/or embryo model.

@@

As an illustrative example, we investigated an enhancer located in one of the introns of \textit{GLI3} (hs111.0; \textbf{Supplementary Fig. S2c}). This region has neural tube specific activity in E11.5 mouse embryos (\textbf{Fig. 2h}). Both

@@

migratory neural crest cell type of the organoid (\textbf{Supplementary Fig. S2d}). Both models predict this region
to be specific to the migratory neural crest class (\textbf{Supplementary Fig. S2e}) and nucleotide importance scores

@@

To investigate the enhancer code underlying dorso-ventral patterning of neuronal progenitors further, we calculated
contribution scores on the top 1,000 regions of each organoid and embryo dorso-ventral topic using respectively $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$. We then used TF-MoDISco\cite{shrikumarTechnicalNoteTranscription2018} to find reoccurring seqlets for each topic (patterns) and we clustered patterns based on pairwise motif similarities. Next, we calculated the average contribution for all identified instances per cluster and for each topic resulting in a dorso-ventral enhancer code-table (\textbf{Fig. 3e}). Both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns
underlying the different dorso-ventral states and the importance of the instances of these patterns is highly correlated across both
models (0.75 average Pearson correlation; \textbf{Fig. 3e}). For the most ventral state (organoid topic 33 and embryo topic 34) a combination of FOX, TEAD
and RFX instances is important (\textbf{Fig. 3e-f}) and the importance of these instances correlates with the expression of respectively \textit{FOXA1/A2/P2/P4}, \textit{TEAD1} and \textit{RFX3/4} in both the organoid and embryo cells (\textbf{Fig. 3g} and Supplementary Fig. S3). In subsequent more dorsal neuronal progenitor domains we observe importance of NKX instances, correlating with \textit{NKX2-1/2} expression (\textbf{(Fig. 3e- g)}); PAX instances, correlating with \textit{PAX3/6} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3) and ZIC instances, correlating with \textit{ZIC1} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). For all dorso-ventral progenitor domains, except the most ventral one (and the most dorsal one in the embryo), SOX instances are important which correlates with \textit{SOX2/3} expression (\textbf{Fig. 3e-g} and Supplementary Fig. S3). Similarly, ZEB instances have a negative importance to the accessibility of regions in dorso-ventral domains except the most ventral one (\textbf{Fig. 3e} and Supplementary Fig. S3). Finally, we calculated the Jaccard index of genomic regions based on the presence of an instance of each identified pattern to investigate which patterns often co-occur in the same regions (\textbf{Fig. 3h}). FOX, TEAD and RFX instances often co-occur in the same genomic region while instances of other patterns seem to be present in separate regions (\textbf{Fig. 3h}).\par

It has been reported that FOXA1/2 is a pioneer factor (REF). Indeed, when centering on FOX instances in both the organoid and
embryo data we observe a strong footprint when looking at Tn5 cut-sites (\textbf{Fig. 3i}). We then
asked if there is any difference between regions that are generally bound by FOXA2 in any cell state that expresses this
TF and regions, bound by FOXA2, that are specific to the floorplate. For this purpose, we downloaded FOXA2 ChIP-seq
peaks in HepG2 and A549 and intersected these with floorplate peaks containing FOX instances. Those floorplate peaks
that overlapped with both ChIP-seq datasets we classified as regions that are generally bound by FOXA2 and those that
did not we classified as floorplate specific. Next, we quantified both the number of FOX instances per region and the
affinity of each individual instance (making use of the FOX motif score calculated using
FIMO\cite{grantFIMOScanningOccurrences2011, schreiberTangermemeBiologicalSequence2025}) hypothesizing that the overall affinity of regions that are generally
bound by FOXA2 is higher than floorplate specific regions either because these general regions have more instances or
the affinity of individual instances is higher. Indeed, FOXA2 target regions that are specific to the floorplate mostly
have a single FOX instance while FOXA2 target regions that are general tend to have more instances (\textbf{Fig. 3j}).
Furthermore, in case there is but a single FOX instance in a general FOXA2 target region the affinity of this instance tends
to be higher (\textbf{Fig. 3k}). Generating a \textit{de novo} motif from general and specific FOXA2 target regions that
have a single instance, we observe that FOX instances in regions that are general tend to be of the form
\textit{TGTTT\underline{AC}} while those instances in specific regions tend to be of the form \textit{TGTTT} (\textbf{Fig. 3l}).\par

@@

$DeepNeuralTube_{embryo}$ was able to distinguish the different AP domains but grouped topics of similar domains
together (i.e., Topic 59, 31 and 62; Topic 52 and 71), resulting in four distinct AP domains (\textbf{Fig.4 c-d}). By
explaining the predictions of $DeepNeuralTube_{embryo}$ on the regions of those domains we found four distinct
homeodomain TF binding motifs, one for each domain (\textbf{Fig. 4e}). From anterior to posterior this is \textit{TAATTA}, \textit{GGATTA}, \textit{TCATGWTGANTGA} and \textit{CATYMATCA}. Notably, the patterns corresponding to the two posterior most topics correspond to dimer-motifs. Previously it was already hypothesized that genome-wide specific binding of homeodomain TFs is accomplished through cooperative binding of two TFs\cite{mannChapter3Hox2009}. Indeed, the presence of instances of the \textit{TCATGWTGANTGA} motif correlates with the expression of both \textit{PAX5/8} and \textit{EN1/2} and the prescence of instances of the \textit{CATYMATCA} motif correlates with the expression of \textit{MEIS1/2}, \textit{PBX1/2} and \textit{HOXB3} (\textbf{Fig. 4d-e} and Supplementary Fig. S4). In contrast, the patterns corresponding to the two anterior most topics correspond to monomer motifs. The presence of instances of these patterns correlates with the expression of \textit{LHX2/9}, \textit{RAX} and \textit{SIX3/6} for the most anterior topic and with the expression of \textit{OTX1/2}, \textit{EMX2}, \textit{DMBX1} and \textit{LMX1A} for the second most anterior topic (\textbf{Fig. 4d-e} and Supplementary Fig. S4).

@@

Focusing on neural crest cells, we identified topics representing cells and regions of both pre- and migratory neural
crest as well as facial mesenchyme (\textbf{Fig. 5a}). Similar to the neuronal progenitors, we used
$DeepNeuralTube_{oranoid}$ and $DeepNeuralTube_{embryo}$ to explain the top 1,000 regions for each organoid and embryo
topic and generated a code-table based on the identified pattern instances (\textbf{Fig. 5b}). Again, both
$DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ found similar patterns and the importance of the instances fo
these patterns is highly correlated across both models (0.73 average Pearson correlation coefficient; \textbf{Fig. 5b}).\par

Organoid Topic 62 and embryo Topic 103 represent the pre-migratory neural crest state and instances of TEAD, GRHL, AP2 and ZEB patterns are important for this state (\textbf{Fig. 5b-c}) the importance of TEAD instances correlates with \textit{TEAD3/4} expression, GRHL with \textit{GRHL1/1} and AP2 with both \textit{TFAP2A} and \textit{TFAP2C} (\textbf{Fig. 5d} and Supplementary Fig. S5). Note that ZEB instances have a positive contribution to the pre-migratory while it has a negative contribution to all other neural crest states (\textbf{Fig. 5b-c}) and moreover all other states of both the neuronal progenitors (\textbf{Fig. 3e}) and neurons (\textbf{Fig. 6b}). The positive contribution of ZEB in the pre-migratory neural crest state coincides with the absence of the expression of both \textit{ZEB1} and \textit{ZEB2} (\textbf{Fig. 5d} and Supplementary Fig. S5). This is in line with the previously reported role of ZEB2 as a chromatin closing repressor \cite{taskiranCelltypedirectedDesignSynthetic2024}.\par

The migratory neural crest cell state is represented by two organoid topics (Topic 65 and 59) and a single embryo topic (Topic 94). Nuclear receptor, AP2, ZIC (only in organoid topics) and SOX-dimer (organoid Topic 59 but not 65) are important for this state (\textbf{Fig. 5b-c}). We could not link the nuclear receptor instances to a specific TF based on the correlation of expression and the importance score of these instances. The importance of AP2 instances correlates with the expression of \textit{TFAP2A}, \textit{TFAP2B} and \textit{TFAP2C} (\textbf{Fig. 5d}), ZIC with \textit{ZIC1} to \textit{ZIC5} and the SOX-dimer with expression of \textit{SOX10} (\textbf{Fig. 5d} and Supplementary Fig. S5). The switch from TFAP2A/C in pre-migratory neural crest to TFAP2A/B in migratory neural crest is consistent with a previously reported function of both heterodimers in respectively neural plate border induction and neural crest specification\cite{rothsteinHeterodimerizationTFAP2Pioneer2020}. After the pre-migratory and before the migratory neural crest state another topic is found in both organoid (Topic 60) and embryo (Topic 105) for which instances of RFX are important (\textbf{Fig. 5b-c}) that correlate with the expression of \textit{RFX3/4} (\textbf{Fig. 5d} and Supplementary Fig. S5).\par

Finally, the facial mesenchyme cell state is represented by organoid Topic 58 and embryo Topic 91. Instances of a FOX
pattern and a dimer consisting of a homeobox and ebox motif are important for this state (\textbf{Fig. 5b}). The
importance of the FOX instances correlates with the expression of \textit{FOXC2/D1/P1/P2} and that of the dimer motif with the
expression of both \textit{TWIST1} and \textit{ALX1/4} and \textit{PRRX1} (\textbf{Fig. 5d} and Supplementary Fig. S5). The latter is consistent with a recent study where it has been
shown that TWIST1 and homeodomain TFs cooperatively bind this specific dimer motif regulating the expression of genes involved in
facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024}.\par

@@

Recently, a genome wide association study (GWAS) was published identifying 203 single-nucleotide polymorphisms (SNPs)
that are associated with facial variation\cite{whiteInsightsGeneticArchitecture2021}. Given that regions surrounding these SNPs were
enriched for enhancer activity in cranial neural crest cells\cite{whiteInsightsGeneticArchitecture2021} we asked if, using
$DeepNeuralTube$, we could explain the effect of some of these variants on enhancer activity. Therefore for all
suggestive SNPs we scored both the reference and alternative allele(s) using both $DeepNeuralTube_{organoid}$ and $DeepNeuralTube_{embryo}$ and calculated the difference in prediction score across both alleles (delta prediction score). At an absolute delta prediction score threshold of 0.2 we identified 235 SNPs that we deem explainable by either of the two models. Note that only very few (1 out of 203) of these "explainable" SNPs overlap with the 203 lead SNPs identified by White \textit{et al.} (\textbf{Fig. 5f}). Most of the "explainable" SNPs have a high delta prediction score for either organoid neural crest Topic 58 or Topic 59 (\textbf{Supplementary Fig. S6}). As an example we show the explanation of rs1555067 which is located upstream of \textit{PRRX1}. The G \textgreater A substitution of this SNP causes a delta prediction score of 0.64 for organoid Topic 58 (\textbf{Fig. 5g}), increasing the predicted chromatin accessibility of this genomic region in the facial mesenchyme cell state. This predicted increase in chromatin accessibility is caused by generating an EBOX site 6bp downstream of a partial homeobox site potentially creating a new binding site for TWIST1-homeodomain TF dimers (\textbf{Fig. 5h}).\par

@@

For both the organoid and embryo we found two topics (organoid Topic 6 and 4 and embryo Topic 10 and 18) for which
instances of patterns from neuronal progenitors, more precisely the floorplate, are important (\textbf{Fig. 6b-c}).
These include instances of TEAD, FOX and RFX (\textbf{Fig. 6b-c}) correlating with the expression of respectively \textit{TEAD1},
\textit{FOXA2/J3/O1/P1/P2/P4} and \textit{RFX3/4} (\textbf{Fig. 6d} and Supplementary Fig. S7). Instances of a SOX monomer motif is important for one of these topics (\textbf{Fig. 6b-c}), correlating with the expression of \textit{SOX2} (\textbf{Fig. 6d} and Supplementary Fig. S7). Next, in both organoids and embryo cells we identified a topic for which both instances of neuronal progenitor patterns as well as an EBOX pattern is important (organoid Topic 23 and embryo Topic 13 \textbf{Fig. 6b-c}) the importance of this EBOX pattern correlates with the expression of \textit{ASCL1} and \textit{NEUROD4}. For subsequent topics SOX instances are not important anymore (\textbf{Fig. 6b-c}) and we find topics with a where a combination of EBOX and EBF instances are important (organoid and embryo Topic 24; \textbf{Fig. 6b-c}) the latter correlating with the expression of \textit{EBF1-3} (\textbf{Fig. 6d} and Supplementary Fig. S7) and topics where a combination of EBF and ONECUT instances are important (organoid Topic 13 and embryo Topic 18; \textbf{Fig. 6b-c}), the latter correlating with the expression of \textit{ONECUT1-3}. Additionally, in the embryo topics instances of NKX are important in these latter two topics, the importance of these instances is anti-correlated with the expression \textit{NKX2-2} and correlated with the expression of \textit{ISL1} (\textbf{Fig. 6d} and Supplementary Fig. S7). Finally, in the organoids we could identify an additional topic for which instances of GATA are important (\textbf{Fig. 6b-c}) correlating with the expression of \textit{GATA2} (\textbf{Fig. 6d} and Supplementary Fig. S7).\par

@@

The enhancer code remains complicated to study\cite{wassermanAppliedBioinformaticsIdentification2004} especially
compared to the genetic code\cite{kimDecipheringMultiscaleQuantitative2023, deboerHoldOutGenome2024}. Nonetheless, there
are some recurrent patterns that are shared among enhancers of different cell types. In particular, enhancers underlying
DV patterning, early neuronal fate acquisition and neural crest tend to consist out of clusters of 4-5 different
TFBS\cite{kimDecipheringMultiscaleQuantitative2023, longEverChangingLandscapesTranscriptional2016} that are located
within one NDR (+/-147bp). That is to say, cooperative binding of multiple TFs seem to underlie
the cell type specificity of enhancers\cite{lambertHumanTranscriptionFactors2018}. Here, we see most evidence for a
model of indirect cooperativity\cite{longEverChangingLandscapesTranscriptional2016} (i.e., flexible billboard enhancer
model\cite{arnostiTranscriptionalEnhancersIntelligent2005}). However, for multiple cell types we also observed evidence
of direct coooperativity between TFs in the form of protein-protein interactions between two TFs. This was the case for
TWIST1 and homeodomain TFs heterodimers in facial mesenchyme\cite{kimDNAguidedTranscriptionFactor2024} and GRHL1/2 dimers, TFAP2A and TFAP2B/C  heterodimer\cite{rothsteinHeterodimerizationTFAP2Pioneer2020} and SOX10 homodimers in neural crest. For all these TFs we found dimer motifs with a fixed number nucleotides in between the two TFBS instances. Of note, for SOX10 we observed three different dimer motif with either 3, 4 or 5 bps spacing with those instances with 4 bps spacing being the most accessible (\textbf{Supplementary Fig. 8}). Finally, we also observed evidence of pioneer activity, where binding of one TF can facilitate the subsequent binding of other TFs to the same enhancer\cite{longEverChangingLandscapesTranscriptional2016}. This was the case for FOXA2 in the floorplate, consistent with previous findings of FOXA TFs\cite{iwafuchi-doiPioneerTranscriptionFactor2016, iwafuchiGeneNetworkTransitions2020}. Others have hypothesized, however, that pioneer activity is not a qualitative feature of a TF but rather dependent on the affinity of interactions between TF and DNA\cite{hansenTestPioneerFactor2022}. Similarly, we observed that enhancers that are specific to the floorplate contain either less FOXA2 binding sites and/or binding sites of lower affinity. The use of suboptimal TFBS is a reoccurring feature of cell type specific enhancers\cite{farleySuboptimizationDevelopmentalEnhancers2015, crockerChapterTwentySevenSoft2016, limAffinityoptimizingEnhancerVariants2024, naqviTransferLearningReveals2025}.\par

@@

\subsection*{Supplementary figures}

\renewcommand{\thefigure}{S\arabic{figure}}

\setcounter{figure}{0}

\newpage
\begin{figure}[p]
  \includegraphics[width=1.0\linewidth]{figures/supplement/Supplemental_figure_1.png}
  \caption{
    \textbf{Ogranoid cell lines and embryo produce similar gene expression profiles}\\
    \textbf{a}, UMAP based on gene expression profiles of organoid cells colored by cell identity and split by cell
    line. \textbf{b}, scatter plot showing gene expression (of highly variable genes) of the same cell type across cell lines. Number top left
    indicates Pearson correlation coefficient. \textbf{c}, scatter plot showing pairwise gene expression (of highly
    variable genes) across cell types and organoid and embryo. Number top left indicates Pearson correlation
    coefficient. \textbf{d}, d histogram showing distribution of Pearson correlation coefficients of average gene
    expression profile per topic, topics representing
    the same biological state(matching) and other topics (non-matching) across organoid and embryo.
  }
\end{figure}
